---
title: "HW4"
output: html_document
date: "2023-04-05"
author: "Zirui Zhang"
---

```{r setup, message=FALSE}
library(readr)
library(tidyverse)
library(dplyr)
library(genetics)
library(knitr)
library(scales)
library(ggplot2)
```

### QUESTION (1)

#### TRAITS: 1.DBP>80(binary); 2.Mean_BP(continuous); CANDIDATE GENES: 1.akt1; 2.prdx; 3.resistin.

```{r, message=FALSE}
fms = read.csv("./FMS_data.csv")
attach(fms)
```

```{r, chunk_set up traits and genes}
# set up traits
Trait.1 = as.numeric(DBP>80)
Trait.2 = Mean_BP
# set up genes and corresponding columns
NamesAkt1Snps = names(fms)[substr(names(fms),1,4)=="akt1"]
fmsAkt1 = fms[,is.element(names(fms),NamesAkt1Snps)]
NamesRanklSnps = names(fms)[substr(names(fms),1,5)=="rankl"]
fmsRankl = fms[,is.element(names(fms),NamesRanklSnps)]
NamesResistinSnps = names(fms)[substr(names(fms),1,8)=="resistin"]
fmsResistin = fms[,is.element(names(fms),NamesResistinSnps)]
```

### QUESTION (2)

#### (a) Association Analysis:

The pvalues for Trait 1 are stored in df1.

```{r chunk_Trait 1, warning=FALSE}
# write a function to record p-values of chi-sq test
Trait1.Function = function(Geno){
  
  ObsTab = table(Trait.1,Geno)
  return(chisq.test(ObsTab)$p.value)

}
# test of the three genes:
pv.akt1.1 = apply(fmsAkt1,2,Trait1.Function)
pv.rankl.1 = apply(fmsRankl,2,Trait1.Function)
pv.resis.1 = apply(fmsResistin,2,Trait1.Function)
trait1 = c(pv.akt1.1, pv.rankl.1, pv.resis.1)
df1 = data.frame(names = names(trait1), values = trait1)
```

The pvalues for Trait 1 are stored in df3.

```{r chunk_Trait 2, warning=FALSE}
# write a function to record p-values of chi-sq test
Trait2.Function = function(Geno){
  ObsTab = table(Trait.2,Geno)
  return(chisq.test(ObsTab)$p.value)
}
# test of the three genes:
pv.akt1.2 = apply(fmsAkt1,2,Trait2.Function)
pv.rankl.2 = apply(fmsRankl,2,Trait2.Function)
pv.resis.2 = apply(fmsResistin,2,Trait2.Function)
trait2 = c(pv.akt1.2, pv.rankl.2, pv.resis.2)
df3 = data.frame(names = names(trait2), values = trait2)
```

#### (b) Multiple Comparisons Adjustment:

##### Trait 1: 

```{r}
# adjusted p-value for trait 1:
pv.akt1.1.adj = p.adjust(pv.akt1.1, method="BH")
pv.rankl.1.adj = p.adjust(pv.rankl.1, method="BH")
pv.resis.1.adj = p.adjust(pv.resis.1, method="BH")
trait1.ad = c(pv.akt1.1.adj, pv.rankl.1.adj, pv.resis.1.adj)
df2 = data.frame(names = names(trait1.ad), values = trait1.ad)
# kable for trait 1:
trait.1 = 
  merge(df1, df2, by = "names", all.x = TRUE) %>% 
  rename(SNP=names, p.value=values.x, ad.p.value=values.y) 
trait.1$Gene <- c(rep("akt", 24), rep("rankl", 4), rep("resistin", 6))
trait.1 %>%
  relocate(Gene) %>% 
  kable()
```

##### Trait 2: 

```{r}
# adjusted p-value for trait 2:
pv.akt1.2.adj = p.adjust(pv.akt1.2, method="BH")
pv.rankl.2.adj = p.adjust(pv.rankl.2, method="BH")
pv.resis.2.adj = p.adjust(pv.resis.2, method="BH")
trait2.ad = c(pv.akt1.2.adj, pv.rankl.2.adj, pv.resis.2.adj)
df4 = data.frame(names = names(trait2.ad), values = trait2.ad)
# kable for trait 1:
trait.2 = 
  merge(df3, df4, by = "names", all.x = TRUE) %>% 
  rename(SNP=names, p.value=values.x, ad.p.value=values.y) 
trait.2$Gene <- c(rep("akt", 24), rep("rankl", 4), rep("resistin", 6))
trait.2 %>%
  relocate(Gene) %>% 
  kable()
```


#### (c) Population Stratification using PCA:

```{r}
# mutate all snps data into numeric, drop infinite and null values
psd = cbind(fmsAkt1, fmsRankl, fmsResistin) %>% 
  as.data.frame() %>% 
  mutate_if(is.character, as.factor) %>% 
  mutate_if(is.factor, as.numeric) %>% 
  na.omit()
# pca using prcomp
psd.pca = prcomp(psd, center=FALSE, scale=FALSE)
screeplot(psd.pca, type="lines")
```

Here the sreeplot indicates that 2 principle components should be enough, to gain a better view, we plotted PC1 vs PC2, PC1 vs PC3 and PC2 vs PC3. 

```{r}
scores = psd.pca$x
scores %>% head(5)
lamda = percent(psd.pca$sdev^2/sum(psd.pca$sdev^2))
lamda
plot(psd.pca$"x"[, 1], psd.pca$"x"[, 2], xlab = "PC1", ylab = "PC2")
plot(psd.pca$"x"[, 1], psd.pca$"x"[, 3], xlab = "PC1", ylab = "PC3")
plot(psd.pca$"x"[, 2], psd.pca$"x"[, 3], xlab = "PC2", ylab = "PC3")
```


#### (d) Summary:
```{r chunk_descriptive statistics}
# trait.1: whether DBP>80 or not
skimr::skim(Trait.1)
table = table(Trait.1)
barplot(table)

# trait.2: average blood pressure
skimr::skim(Trait.2)
df = as.data.frame(Trait.2)
ggplot(df, aes(x=Trait.2))+
  geom_histogram()
```

